Stage-Splitting a Modal Language

### 2014-10-01 paper/example.tex
fix subfigure references

\section {Example: Staged Quickselect in \lang}
\label{sec:staging}

\begin{figure*}
\label{fig:quickselect}
\begin{minipage}{0.5\textwidth}
\begin{lstlisting} 
datatype list = Empty | Cons of int * list
fun partition (p : int, l : list) 
  : (int*list*list) =
  case unroll l of 
    Empty => (0,Empty, Empty) 
  | Cons (h,t) =>
      let (s,left,right) = partition (p,t) in
      if h<p 
      then (s+1,Cons(h,left),right)
      else (s,left,Cons(h,right))

fun qSelect (l : list, k : int) : int = 
  case l of
    Empty => 0
  | Cons (h,t) => 
      let (left,right,n) = partition h t in
        case compare k n of
          LT => qSelect (left, k)
        | EQ => h
        | GT => qSelect (right, k-n-1)
\end{lstlisting}
\caption{Unstaged implementation of quickselect.}
\end{minipage}
\begin{minipage}{0.5\textwidth}
\begin{lstlisting} 
datatype list = Empty | Cons of int * list
fun partition (p : int, l : list) = ...

fun qsStaged (l : list, k : $int) : $int = 
  case l of
    1`Empty` => next {2`0`}
  | Cons (h,t) => 
      let (left,right,n) = partition h t in
      next{
        let n = hold{n} in
          case compare prev{k} n of
            LT => prev {qsStaged (left k)}
          | EQ => hold {h}
          | GT => prev {qsStaged (right, 
                             next{prev{k}-n-1)}}
      }	
\end{lstlisting}
\caption{Staged implementation of quickselect in \lang.}
\end{minipage}
\begin{minipage}{0.5\textwidth}
\begin{lstlisting}
datatype list = Empty | Cons of int * list
fun partition (p : int, l : list) = ...

@@

fun qSelect2 (p : tree, k : int) : int = 
  case unroll p of
    Leaf => 0
  | Branch (n,h,p1,p2) => 
      case compare k n of
        LT => qSelect2 (p1,k)
      | EQ => h
      | GT => qSelect2 (p2,k-n-1)
\end{lstlisting}
\caption{Split (two-phase) implementation of quickselect.}
\end{minipage}
\caption{Caption place holder}
\end{figure*}

As an example of an algorithm amenable to staging techniques,
consider the implementation of quickselect shown in Figure~2a.
Quickselect is an $O(n)$ (expected time) algorithm for finding the $k$-th largest element in a randomly ordered length-$n$ list.
It does so by partitioning the input list into sublists containing elements less than and greater than the head element, 
then recurring on the sublist containing the desired element.

Importantly, one property of quickselect is that the partitioning of the list, even through its recursive calls, depends only on the values of list elements, not on the value of $k$ 
\footnote{There is a control-dependence on $k$ -- that is $k$ determines what splits are made but not how they are made -- we ignore this for now.}.  Therefore, if quickselect was to be evaluated many times on the same list, this independence admits a familiar optimization where the algorithm is split into two functions: one which precomputes \emph{all partitions} of the input list needed for any value of $k$ (resulting in a binary search tree---see \texttt{qSelect1} in Figure~2c),
and another which, given $k$, uses the tree to perform accelerated lookup of the $k$-th element (essentially performing binary search of a sorted list---see \texttt{qSelect2}).
%One advantage of splitting the algorithm into two phases is that the search tree can be constructed for a list once
%and then used to accelerate many future lookups for any value of $k$.

@@

Figure 2b shows a staged implementation of quickselect in \lang.
As before, the function accepts as input the list to select from and a value of $k$,
but now the latter is represented with the type $\fut\rmint$ rather than $\rmint$.
The difference here is that an $\rmint$ is an integer available at the current stage (stage \bbone), 
whereas a $\fut\rmint$ is an integer available only at the next stage (stage \bbtwo).
Naturally the output type, representing the $k$th largest element of the list,
is also $\fut\rmint$, since it is not computed until stage \bbtwo.

@@

But for the $\fut$, $\next$, $\prev$, and $\pause$ constructs, 
the staged version of quickselect in Figure~2b is virtually identical to the unstaged version in Figure~2a.
The constructs that were added were placed in order to maximize the work done at stage~\bbone\ while still conforming to the type signature.
It would have also been valid to simply move the whole input list unchanged into stage~\bbtwo\ at the very beginning, 
but that would not be particularly interesting since it would result in an effectively trivial split 
that's just the identity at stage~\bbone\ and plus quickselect at stage~\bbtwo.
There has been the extensive research into the question of how to automatically add staging annotations to unstaged code.
This process is known as {\em binding time analysis}, and we do not consider it here.
Instead, we assume that all input programs are properly staged according to some programmer's intent.

>>>>

\section {Example: Staged Quickselect in \texorpdfstring{\lang}{λ12}}
\label{sec:staging}

\begin{figure*}
\begin{subfigure}{0.5\textwidth}
\begin{lstlisting} 
datatype list = Empty | Cons of int * list
fun partition (p : int, l : list) 
  : (int*list*list) =
  case unroll l of 
    Empty => (0,Empty, Empty) 
  | Cons (h,t) =>
      let (s,left,right) = partition (p,t) in
      if h<p 
      then (s+1,Cons(h,left),right)
      else (s,left,Cons(h,right))

fun qSelect (l : list, k : int) : int = 
  case l of
    Empty => 0
  | Cons (h,t) => 
      let (left,right,n) = partition h t in
        case compare k n of
          LT => qSelect (left, k)
        | EQ => h
        | GT => qSelect (right, k-n-1)
\end{lstlisting}
\caption{Unstaged implementation of quickselect.}
\label{fig:quickselect}
\end{subfigure}%
\begin{subfigure}{0.5\textwidth}
\begin{lstlisting} 
datatype list = Empty | Cons of int * list
fun partition (p : int, l : list) = ...

fun qsStaged (l : list, k : $int) : $int = 
  case l of
    1`Empty` => next {2`0`}
  | Cons (h,t) => 
      let (left,right,n) = partition h t in
      next{
        let n = hold{n} in
          case compare prev{k} n of
            LT => prev {qsStaged (left k)}
          | EQ => hold {h}
          | GT => prev {qsStaged (right, 
                             next{prev{k}-n-1)}}
      }	
\end{lstlisting}
\caption{Staged implementation of quickselect in \lang.}
\label{fig:qsstaged}
\end{subfigure}
\begin{subfigure}{0.5\textwidth}
\begin{lstlisting}
datatype list = Empty | Cons of int * list
fun partition (p : int, l : list) = ...

@@

fun qSelect2 (p : tree, k : int) : int = 
  case unroll p of
    Leaf => 0
  | Branch (n,h,p1,p2) => 
      case compare k n of
        LT => qSelect2 (p1,k)
      | EQ => h
      | GT => qSelect2 (p2,k-n-1)
\end{lstlisting}
\caption{Split (two-phase) implementation of quickselect.}
\label{fig:qssplit}
\end{subfigure}
\caption{Caption place holder}
\end{figure*}

As an example of an algorithm amenable to staging techniques,
consider the implementation of quickselect shown in \ref{fig:quickselect}.
Quickselect is an $O(n)$ (expected time) algorithm for finding the $k$-th largest element in a randomly ordered length-$n$ list.
It does so by partitioning the input list into sublists containing elements less than and greater than the head element, 
then recurring on the sublist containing the desired element.

Importantly, one property of quickselect is that the partitioning of the list, even through its recursive calls, depends only on the values of list elements, not on the value of $k$ 
\footnote{There is a control-dependence on $k$ -- that is $k$ determines what splits are made but not how they are made -- we ignore this for now.}.  Therefore, if quickselect was to be evaluated many times on the same list, this independence admits a familiar optimization where the algorithm is split into two functions: one which precomputes \emph{all partitions} of the input list needed for any value of $k$ (resulting in a binary search tree---see \texttt{qSelect1} in \ref{fig:qssplit}),
and another which, given $k$, uses the tree to perform accelerated lookup of the $k$-th element (essentially performing binary search of a sorted list---see \texttt{qSelect2}).
%One advantage of splitting the algorithm into two phases is that the search tree can be constructed for a list once
%and then used to accelerate many future lookups for any value of $k$.

@@

\ref{fig:qsstaged} shows a staged implementation of quickselect in \lang.
As before, the function accepts as input the list to select from and a value of $k$,
but now the latter is represented with the type $\fut\rmint$ rather than $\rmint$.
The difference here is that an $\rmint$ is an integer available at the current stage (stage \bbone), 
whereas a $\fut\rmint$ is an integer available only at the next stage (stage \bbtwo).
Naturally the output type, representing the $k$th largest element of the list,
is also $\fut\rmint$, since it is not computed until stage \bbtwo.

@@

But for the $\fut$, $\next$, $\prev$, and $\pause$ constructs, 
the staged version of quickselect in \ref{fig:qsstaged} is virtually identical to the unstaged version in \ref{fig:quickselect}.
The constructs that were added were placed in order to maximize the work done at stage~\bbone\ while still conforming to the type signature.
It would have also been valid to simply move the whole input list unchanged into stage~\bbtwo\ at the very beginning, 
but that would not be particularly interesting since it would result in an effectively trivial split 
that's just the identity at stage~\bbone\ and plus quickselect at stage~\bbtwo.
There has been the extensive research into the question of how to automatically add staging annotations to unstaged code.
This process is known as {\em binding time analysis}, and we do not consider it here.
Instead, we assume that all input programs are properly staged according to some programmer's intent.
